Using Deep Learning to Identify Cognitive Impairment in Electronic Health Records

### 2021-09-08 Submission/Paper/ML4H/Paper.tex
first draft of conclusion -- still short so will iterate more

\section{Results}
\label{sec:Results}  
% talk abt performance
% which patients are correctly classified for each model
% compare conf mat's for each model
% distribution of CI/not CI labels
% talk about different patient aggregation scheme
% address lack of not patient level labels
% TFIDF might have lot of false positives b/c of use of vocab that doesn't discuss patient, deep learning will be able to use context
% start with justification to aggregation to patient level
% sensivity should hopefully improve, deep learning should reduce specificity b/c of use of contextual information
% discuss if certain patients were predicted differently for all models
% some ppl will be classified as 0, until deep learning, where it will be hopefully changed to 1
% group of ppl who get 0 for baseline, prediction of 1 with tf-idf and deep
% group of ppl who get 1 for tf-idf, 0 for deep learning b/c of high false positive rate, vocab cld be talking abt family member (tf-idf will address underdiagnosis, deep learning will address misdiagnosis better) 
% tf-idf is easy to "trigger", deep learning will evaulate based on context
% Talk about Stats
% TF-IDF model struggles with understanding context of words around keyword match
% Future plans include reducing false positives by gathering more annotated sequences and using deep learning natural language. Currently, many of our false positives are from the model identifying the presence of a keyword but struggling to understand the context around the keyword match; many of these contexts negate the presence of CI for the patient in question. Deep Learning can leverage contextual information, making it promising for this task.

We evaulated each model based on sequence level class assignments. Model performance for each model in shown in \tableref{tab:table2}. TF-IDF achieved AUC of 0.95 and accuracy of 0.84. The 20 words with the highest correlation coefficients using TF-IDF word vectorization are shown in Appendix \ref{app:app3}. While TF-IDF was able to identify the presence of a keyword or always pattern in a sequence, it was unable to leverage the context around each keyword match, which occasionally would negate the presence of CI. This led to the model over-zealously predicting sequences as demonstrating CI. ClinicalBERT was able to overcome these issues due to its high complexity transformer architecture, and achieved AUC of 0.98 and accuracy of 0.93. In order to generate accurate patient level class assignments, we applied ClinicalBERT to 186,730 sequences from the N = 13,941 unique patients that were not patient of our training/validation/set sets. With these sequence level predictions, we generated patient level class assignments by empirically tuning a sequence threshold that represented the number of sequences predicted as Positive by ClinicalBERT. We identified the most optimal threshold by comparing the percentage of patients being predicted as having CI (stratified by APOE allele) to the percentages of patients with CI related Meds/ICD codes (also stratified by APOE allele). \tableref{tab:table4} and \figureref{fig:fig2} shows the comparison of Med/ICD codes to ClinicalBERT patient level class assignments with a sequence threshold of 2. As shown, ClinicalBERT was able to identify a significant proportion of patients that went undetected by current clinical methods, highlighting the utility of such a tool in a clinical environment.

\begin{table*}[hbtp]
\floatconts
{tab:table4}
\centering 
    {\begin{tabular}{cccccc}
    \toprule 
    \bfseries  & \bfseries Count & \bfseries Yes (\%) & \bfseries No/Ntr (\%)  & \bfseries Med/ICD Code (\%) \\ 
    \midrule
    \bfseries{APOE ${\bm{\varepsilon}}$2} & \fseries 1754 & \fseries 0.17 & \fseries 0.83 & \fseries 0.11  \\
    \bfseries{APOE ${\bm{\varepsilon}}$3} & \fseries 8751 & \fseries 0.17 & \fseries 0.83 & \fseries 0.11  \\
    \bfseries{APOE ${\bm{\varepsilon}}$4} & \fseries 3436 & \fseries 0.21 & \fseries 0.79 & \fseries 0.17  \\

    \bottomrule
  \end{tabular}}
  {\caption{Comparison between Other Indicators of Cognitive Impairment and Clinical-BERT}} \\
\end{table*}

\begin{figure*}[h!]
\label{fig:fig2}
\centering 
\includegraphics[scale = 0.75]{bert_patient_predictions.png}
\caption{Patient Level Predictions for ClinicalBERT}
\end{figure*}

%% Conclusion + Future Work %%
\section{Conclusion and Future Work} 
\label{sec:ConclusionFutureWork} We applied innovative Machine Learning and Deep Learning NLP techniques to identify patients with cognitive impairment in EHR. Our work can help combat the issues of underdiagnosis and misdiagnosis for dementia. % {More coming soon } %

\clearpage
\appendix

@@

@@

% Appendix B
\section{Keywords} 
\begin{table}[hbtp] 
\floatconts
{app:keywords}
    {\begin{tabular}{lccc}
        \toprule
        \bfseries Keyword & \bfseries Match Count\\
        \midrule

        \bottomrule
        \end{tabular}}
        {\caption{Keywords indicative of Cognitive Impairment}} \\

\end{table}

@@

% Figure 1 - Example Sequences
\section{Example Sequences} 
\begin{figure}[h!] \label{app:examples}
\centering 
\includegraphics[scale = 0.35]{Figure-1.png}
\caption{Example Sequences}
\end{figure}
\clearpage

% Appendix C
\section{Top TF-IDF Word Features} \label{app:app3}
\begin{table}[hbtp]
    {\begin{tabular}{cccc}
    \toprule 
    \bfseries Word & \bfseries Corr & \bfseries Word & \bfseries Corr &
    \midrule

@@

    \bottomrule
  \end{tabular}}
\end{table}
\clearpage

>>>>

\section{Results}
\label{sec:Results}  
% talk abt performance
% which patients are correctly classified for each model
% compare conf mat's for each model
% distribution of CI/not cognitive impairmentlabels
% talk about different patient aggregation scheme
% address lack of not patient level labels
% TFIDF might have lot of false positives b/c of use of vocab that doesn't discuss patient, deep learning will be able to use context
% start with justification to aggregation to patient level
% sensivity should hopefully improve, deep learning should reduce specificity b/c of use of contextual information
% discuss if certain patients were predicted differently for all models
% some ppl will be classified as 0, until deep learning, where it will be hopefully changed to 1
% group of ppl who get 0 for baseline, prediction of 1 with tf-idf and deep
% group of ppl who get 1 for tf-idf, 0 for deep learning b/c of high false positive rate, vocab cld be talking abt family member (tf-idf will address underdiagnosis, deep learning will address misdiagnosis better) 
% tf-idf is easy to "trigger", deep learning will evaulate based on context
% Talk about Stats
% TF-IDF model struggles with understanding context of words around keyword match
% Future plans include reducing false positives by gathering more annotated sequences and using deep learning natural language. Currently, many of our false positives are from the model identifying the presence of a keyword but struggling to understand the context around the keyword match; many of these contexts negate the presence of cognitive impairmentfor the patient in question. Deep Learning can leverage contextual information, making it promising for this task.

We evaluated each model based on sequence level class assignments. Model performance for each model in shown in \tableref{tab:table2}. TF-IDF achieved AUC of 0.95 and accuracy of 0.84. The 20 words with the highest correlation coefficients using TF-IDF word vectorization are shown in Appendix \ref{app:app3}. While TF-IDF was able to identify the presence of a keyword or always pattern in a sequence, it was unable to leverage the context around each keyword match, which occasionally would negate the presence of CI. This led to the model over-zealously predicting sequences as demonstrating CI. ClinicalBERT was able to overcome these issues due to its high complexity transformer architecture, and achieved AUC of 0.98 and accuracy of 0.93. In order to generate accurate patient level class assignments, we applied ClinicalBERT to 186,730 sequences from the N = 13,941 unique patients that were not patient of our training/validation/set sets. With these sequence level predictions, we generated patient level class assignments by empirically tuning a sequence threshold that represented the number of sequences predicted as Positive by ClinicalBERT. We identified the most optimal threshold by comparing the percentage of patients being predicted as having cognitive impairment(stratified by APOE allele) to the percentages of patients with cognitive impairment related Meds/ICD codes (also stratified by APOE allele). Appendix \ref{app:bert} shows the comparison of Med/ICD codes to ClinicalBERT patient level class assignments with a sequence threshold of 2. As shown, ClinicalBERT was able to identify a significant proportion of patients that went undetected by current clinical methods, highlighting the utility of such a tool in a clinical environment.

%% Conclusion + Future Work %%
\section{Conclusion and Future Work} 
\label{sec:ConclusionFutureWork} We applied NLP algorithms to identify patients with cognitive impairment in EHR and
compared a machine learning model with a deep learning model on performance of sequence level class assignment predictions. Our work can help combat the issues of under-diagnosis and misdiagnosis for dementia. The the deep learning model's performance was significantly better than the machine learning model. We used the sequence level class assignments of the deep learning model to generate patient level class assignment by assigning a patient a label of cognitive impairment if their number of sequences predicted positive was greater than or equal to a empirically tuned threshold. However, a lack of patient level class assignments generated by annotators prevents us from measuring the true accuracy of our results. In order to fix this issue, we plan to gather 1000 patient level class assignments. We also plan to further improve the generalisability of our models by gathering more manually reviewed sequences. An active learning loop will be used to pick particular patients and sequences by using entropy score to label uncertain cases and UMAP clustering of ClinicalBERT word embeddings.

\appendix

@@

% Appendix C
\section{Keywords} 
\begin{table}[hbtp] 
\floatconts
{app:keywords}
    {\begin{tabular}{lccc}
        \toprule
        \bfseries Keyword & \bfseries Match Count\\
        \midrule

        \bottomrule
        \end{tabular}}
        {\caption{Keywords indicative of Cognitive Impairment}} \\
\end{table}

@@

% Appendix D / Figure 1 - Example Sequences
\section{Example Sequences} 
\begin{figure}[h!] \label{app:examples}
\centering 
\includegraphics[scale = 0.35]{Figure-1.png}
\caption{Example Sequences}
\end{figure}

@@

% Appendix E
\section{Top TF-IDF Word Features} \label{app:app3}
\begin{table}[hbtp]
    {\begin{tabular}{cccc}
    \toprule 
    \bfseries Word & \bfseries Corr & \bfseries Word & \bfseries Corr &
    \midrule

@@

    \bottomrule
  \end{tabular}}
\end{table}

\section{Assessment of ClinicalBERT Patient Level Class Assignments} \label{app:bert}

\begin{table}[hbtp]
\floatconts
{tab:table4}
\centering 
    {\begin{tabular}{cccccc}
    \toprule 
    \bfseries  & \bfseries Count & \bfseries Yes (\%) & \bfseries No/Ntr (\%)  & \bfseries Med/ICD Code (\%) \\ 
    \midrule
    \bfseries{APOE ${\bm{\varepsilon}}$2} & \fseries 1754 & \fseries 0.17 & \fseries 0.83 & \fseries 0.11  \\
    \bfseries{APOE ${\bm{\varepsilon}}$3} & \fseries 8751 & \fseries 0.17 & \fseries 0.83 & \fseries 0.11  \\
    \bfseries{APOE ${\bm{\varepsilon}}$4} & \fseries 3436 & \fseries 0.21 & \fseries 0.79 & \fseries 0.17  \\

    \bottomrule
  \end{tabular}}
  {\caption{Comparison between Other Indicators of Cognitive Impairment and Clinical-BERT}} \\
\end{table}

\begin{figure}[h!]
\label{fig:fig2}
\centering 
\includegraphics[scale = 0.75]{bert_patient_predictions.png}
\caption{Confusion Matrix Patient Level Prediction Counts for ClinicalBERT}
\end{figure}
